Retrofitting Effect Handlers onto OCaml

### 2020-11-12 retro-concurrency.tex
Updates

\vspace{-2mm}
\section{Semantics}
\label{sec:semantics}

@@

\subsection{Dynamic semantics}

@@

\subsection{External calls and callbacks}

@@

\subsection{Stack unwinding}

@@

One challenge is to encode in DWARF directives the CFA when stacks are switched
at effect handlers, callbacks and external calls. When starting to run on a new
fiber, we insert DWARF bytecode to follow the |parent_fiber| pointer and
dereference the |saved_sp| to get the CFA (|saved_sp + 8|). During callbacks
into OCaml, we save the current system stack pointer in the |DWARF block| in
Figure~\ref{fig:fiber} to identify the CFA in the C stack. DWARF unwinding for
external calls is implemented by following a link to the current OCaml stack
pointer. With these changes, we get the same backtrace for the \texttt{meander}
program from \S\ref{sec:unwind}, modulo runtime system functions due to effect
handlers. We have utilized a DWARF verification tool~\cite{Bastian19} with
extensions to handle PIE and additional DWARF primitives to flag possible errors
in the hand written DWARF directives where we have altered the OCaml assembler
to implement effects. In three places additional manual verification was
required; two were where the stack is intentionally switched between fibers and
one was where the return to the parent stack on child stack completion could not
track stack resizes and intermediate effect handler frames.

\subsection{Garbage collection}

@@

\section{Evaluation}
\label{sec:eval}

@@

\subsection{No effects benchmarks}

@@

\subsection{No perform benchmarks}

@@

\subsection{Concurrent benchmarks}

@@

\section{Related Work}
\label{sec:related}

@@

We presented the motivation, design and implementation of effect handlers
specifically aimed at retrofitting them onto OCaml with tool compatibility,
negligible overheads for existing OCaml programs and good performance for those
that use effect handlers. Our evaluation shows that this is indeed the case. We
plan to upstream the compiler and runtime support for effect handlers followed
by an effect system.

@@

>>>>

\section{Semantics}
\label{sec:semantics}

@@

\vspace{-5mm}
\subsection{Dynamic semantics}

@@

\vspace{-3mm}
\subsection{External calls and callbacks}

@@

\vspace{-2mm}
\subsection{Stack unwinding}

@@

One challenge is to encode in DWARF directives the CFA when stacks are switched
at effect handlers, callbacks and external calls. When starting to run on a new
fiber, we insert DWARF bytecode to follow the |parent_fiber| pointer and
dereference the |saved_sp| to get the CFA (|saved_sp + 8|). During callbacks
into OCaml, we save the current system stack pointer in the |DWARF block| in
Figure~\ref{fig:fiber} to identify the CFA in the C stack. DWARF unwinding for
external calls is implemented by following a link to the current OCaml stack
pointer. With these changes, we get the same backtrace for the \texttt{meander}
program from \S\ref{sec:unwind}, modulo runtime system functions due to effect
handlers.

We have utilized a DWARF verification tool~\cite{Bastian19} with extensions to
handle PIE and additional DWARF primitives to flag possible errors in the hand
written DWARF directives where we have altered the OCaml assembler to implement
effects. In three places additional manual verification was required; two were
where the stack is intentionally switched between fibers and one was where the
return to the parent stack on child stack completion could not track stack
resizes and intermediate effect handler frames.

Even with correct DWARF unwind information, using DWARF to record call stack
information in |perf| only captures the call stack of the current fiber in
Multicore OCaml. Since stack unwinding using DWARF is slow due to bytecode
interpretation overhead, |perf| dumps the (user) call stack when
sampled~\cite{Bastian19}. This only includes the frames from the current fiber.
This is a limitation of |perf| and not of our stack layout.

\vspace{-2mm}
\subsection{Garbage collection}

@@

\vspace{-3mm}
\section{Evaluation}
\label{sec:eval}

@@

\vspace{-3mm}
\subsection{No effects benchmarks}

@@

\vspace{-3mm}
\subsection{No perform benchmarks}

@@

\vspace{-3mm}
\subsection{Concurrent benchmarks}

@@

\vspace{-3mm}
\section{Related Work}
\label{sec:related}

@@

We presented the motivation, design and implementation of effect handlers
specifically aimed at retrofitting them onto OCaml with tool compatibility,
negligible overheads for existing OCaml programs and good performance for those
that use effect handlers. Our evaluation shows that this is indeed the case.

@@

In addition, multi-shot continuations currently are not safe under compiler
optimisations. For example consider the following program:
\begin{lstlisting}
effect E : unit

let foo () = perform E

let bar f =
  let r = ref 0 in
	f ();
  r := !r + 1;
  !r

let _ = match bar foo with
| v -> v
| effect E k ->
    let k' = Obj.clone_continuation k in
    let r1 = continue k () in
    let r2 = continue k' () in
    Printf.printf "%d %d\n" r1 r2;
		0
\end{lstlisting}
This program is expected to print |1 2|. However, OCaml optimizes |bar| by
converting the heap allocation of |r| to a stack allocation. As a result, each
continuation ends up with a copy of |r|, and the program prints |1 1|. In the
presence of multi-shot continuations, compiler optimisations need to be made
aware of the effects of a function. We leave proper support for multi-shot
continuations to future work.
